A statistical framework for analysing whole-genome bisulfite-sequencing data

### 2015-05-03 Rmarkdown/wgbs_statistical_framework.Rmd
Edits to WGBS statistical framework

---
title: "A statistical framework for whole-genome bisulfite-sequencing data"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\chapter{A statistical framework for whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}

@@

The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state. As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile. Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states. Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.

@@

From these we can compute the proportion of fragments that methylated at the $i^{th}$ locus:

@@

Again, I emphasise that $H_{(i, i + 1, i + \text{m} - 1), j}, Z_{h, (i, i + 1, i + \text{m} - 1), j}, B_{(i, i + 1, i + \text{m} - 1), j}$ and the set of methylation patterns are unobservable. However, by sequencing the pools of DNA fragments we aim to estimate these variables.

@@

In a whole-genome bisulfite-sequencing experiment we do not sequence every DNA fragment in the pool. Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments. We have a large number (~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments. Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to the number of fragmentation containing the $i^{th}$ methylation locus, $H_{i}$.

@@

1. There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4))$. In effect, we have missing data for any intervening methylation loci, the $(i + 2)^{th}$ loci in this example.
2. Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g. _read-1_ gives us $(i, i + 1)$ and _read-2_ gives us $(i + 1, i + 2, i + 3)$. In this example we must use only one of _read-1_ or _read-2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.

@@

Again, these definitions can be extended from 1-tuples to m-tuples. The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$ and has sequencing depth $d_{(i, i + 1, \ldots, i + \text{m} - 1)} = |\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}|$. A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:

@@

Note that we do not know from which DNA fragment, $h$, each read came from, only that all methylation loci in the read came from the same DNA fragment.

@@

The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}}$. In this section I review various methods for estimating these key parameters. When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable. I have suppressed $j$ subscript when referring to a single sample.

@@

Recently, more sophisticated methods have been proposed to estimate or model the average methylation level. These methods, which are still based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.

@@

Several papers have proposed the beta-binomial distribution since, as noted by \citet{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" . The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing. Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing. This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data. For example, the `edgeR` software \citep{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.

@@

`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$. Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{d_{i, j}}$. Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample. This phenomenon, called _co-methylation_ is discussed and analysed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}).

@@

`bsseq` assumes for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$. In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.

@@

\citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`. Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`. However, these comparison studies use RRBS data, which `bsseq` is not designed for`, and so will favour methods designed for RRBS data[^rrbs_sim].

[^rrbs_sim]: Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} altered the default `bsseq` parameters to try to make them comparable to `BiSeq`. \citet{Hebestreit:2013ko} changed the default minimum window size to 80bp but still required at least 20 CpGs per window. \citet{Lacey:2013iy} kept the default minimum window size of 2000 bp but reduced the minimum number of CpGs per window to 50 from the default of 70. Nevertheless, the fact remains that `bsseq` is designed for analysing whole-genome bisulfite-sequencing data and not RRBS, which puts it at a disadvantage in these comparisons.

@@

Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{n_{i, j}}}$ \citep{Hansen:2012gr}. The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample. In this section I discuss this interpretation and statistical properties of this estimator.

@@

What is immediately clear is that these distributions are bimodal: most CpGs are highly methylated or lowly methylated. The exception is the _E16.5\_male\_1_ sample from the _Seisenberger_ data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \cite{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.

@@

These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands. While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample, except for the _E16.5\_male\_1_ sample. The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated. These plots also show that most of the intermediate methylation occurs outside of the CpG islands.

@@

Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5 -- 0.7$. The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$). This suggests that it may not be a good idea to combine CpG methylation levels across strands for embryonic stem cell samples.

@@

The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine. Note that this refers to individual cytosines, $Z_{h, i}$, and not the genomic position of the locus, $pos_{i}$. Also recall that most assays measure DNA methylation from a pool of cells, not a single haploid copy of the genome. The methylation state at the $i^{th}$ locus may vary across the $\mathcal{H}_{i}$ DNA fragments. Therefore, I do not think it makes sense to describe a locus, $i$, as being a 'methylcytosine'. However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have used this latter definition, which I believe to be an unnecessary source of confusion.

@@

Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, nor is any code made available, I believe the method is as follows[^mathematical_description]. For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = d_{i}} Pr(X = k)$, where $X = Binom(d_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'. Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]

@@

\citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values for conducting differential methylation analysis from microarray data[^interpretation]. The main reason they advocate for the use of $\mathcal{M}$-values is that they are approximately _homoscedastic_, i.e. their variances are approximately constant across the full range of $\mathcal{M}$-values. As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformaticians and genomics researchers makes it a favourable choice.

@@

As with $\beta$-values, $\mathcal{M}$-values derived from bisulfite-sequencing data generally cannot be directly analysed due to the variable sequencing coverage across loci. However, the variable sequencing coverage can be accounted for using the afore-mentioned beta-binomial models and related regression models, which are discussed in the next section.

@@

In this chapter I have defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. I have sought to address some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. By using a common statistical framework we can better understand how different statistical methods relate to one another. Using this framework, I have described common estimators of DNA methylation levels and some of their statistical properties. I then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.

Using these data, I have shown that the CpG islands drive the strong bimodal distribution of $\beta$-values that is seen in almost all samples and that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_). The genome-level results for the _Seisenberger_ samples are more difficult to interpret, however, since they are from pooled DNA. Nonetheless, some of the increased variability in the _Seisenberger_ data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.

>>>>

---
title: "A statistical framework for analysing whole-genome bisulfite-sequencing data"
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: ../latex/header.tex
  html_document:
    keep_md: true
bibliography: ../latex/phd_thesis.bib
---

\chapter{A statistical framework for analysing whole-genome bisulfite-sequencing data}\label{chap:wgbs_statistical_framework}

@@

The methylation state of a locus can vary within a sample due to cell-to-cell variability of DNA methylation. A sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell may have a slightly different methylation profile. Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states. Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.

@@

From these we can compute the proportion of fragments that are methylated at the $i^{th}$ locus:

@@

Again, I emphasise that $H_{(i, i + 1, i + \text{m} - 1), j}, Z_{h, (i, i + 1, i + \text{m} - 1), j}, B_{(i, i + 1, i + \text{m} - 1), j}$ and the set of methylation patterns are unobservable. We aim to estimate these variables through sequencing the pool of DNA fragments.

@@

A whole-genome bisulfite-sequencing experiment does not sequence every DNA fragment in the pool. Rather, sequencing can be thought of as sampling without replacement from the pool of DNA fragments. We have a large number (~$10^{10}$) of fragments in the pool and each methylation locus is only present on a small number of those fragments. Therefore, we can approximate this sampling by Poisson sampling, where the rate parameter for locus $i$ is proportional to the number of fragments in the pool and inversely proportional to the number of fragmentation containing the $i^{th}$ methylation locus, $H_{i}$.

@@

1. There may be gaps due to the insert size being longer than the sum of the read lengths, e.g., $(i, i + 1, i + 3, i + 4)$. In effect, we have missing data for any intervening methylation loci, the $(i + 2)^{th}$ loci in this example.
2. Loci may be measured twice if the insert size is less than the sum of the read lengths, e.g. _read-1_ gives us $(i, i + 1)$ and _read-2_ gives us $(i + 1, i + 2, i + 3)$. In this example we must use only one of _read-1_ or _read-2_ as the measurement of the $(i + 1)^{th}$ locus because otherwise we are 'double-counting'.

@@

These definitions can also be extended from 1-tuples to m-tuples. The set of all reads containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$ and has sequencing depth $d_{(i, i + 1, \ldots, i + \text{m} - 1)} = |\mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}|$. A single read containing the m-tuple $(i, i + 1, \ldots, i + \text{m} - 1)$ is denoted by $z: z \in \mathcal{R}_{(i, i + 1, \ldots, i + \text{m} - 1)}$, and the observed methylation state is given by:

@@

Note that we do not know from which DNA fragment ($h$) each read came from, only that all methylation loci in the read came from the same DNA fragment.

@@

The main parameter of interest for each sample is the vector of methylation levels for each locus, $\bm{B_{j}} = (B_{1,j}, \ldots, B_{N_{loci}, j})$. In this section I review various methods for estimating these key parameters. When necessary, I have 'translated' the original work into my notation to make these methods more readily comparable. I have suppressed $j$ subscript when referring to a single sample.

@@

More sophisticated methods have recently been proposed to estimate or model the average methylation level. These methods, which are still based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citep{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \citep{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.

@@

Several papers have proposed the beta-binomial distribution since, as noted by \citet{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" . The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing. Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing. This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data. For example, the `edgeR` software \citep{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-Poisson mixture distribution, to account for both the biological and sampling variability.

@@

`BSmooth`, published in \citet{Hansen:2011gu, Hansen:2012gr} and available in the R/Bioconductor package `bsseq`, and `BiSeq`, published in \citet{Hebestreit:2013ko} and available in the R/Bioconductor package `BiSeq`, take a different approach to getting improved estimates of the $B_{i, j}$. Both `bsseq` and `BiSeq` use statistical smoothing of the 'raw' $\beta_{i, j} = \frac{m_{i, j}}{d_{i, j}}$. Smoothing is motivated and justified by the fact that the $B_{i, j}$ are spatially correlated within a sample. This phenomenon, called _co-methylation_ is discussed and analysed in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.

@@

`bsseq` assumes that for each sample that the underlying methylation level, $B_{i, j}$, is a smoothly varying function of the position in the genome, $i$. In contrast, `BiSeq` first creates clusters of CpGs and only assumes that the underlying methylation level is smooth at positions within each cluster.

@@

\citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} compare the smoothing results of `BiSeq` to `bsseq`. Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} provide instances where they claim `BiSeq` gives more 'reasonable' smoothed values than `bsseq`. However, these comparison studies use RRBS data, which `bsseq` is not designed for`, and their simulation so will favour methods designed for RRBS data[^rrbs_sim].

[^rrbs_sim]: Both \citet{Hebestreit:2013ko} and \citet{Lacey:2013iy} altered the default `bsseq` parameters to try to make them comparable to `BiSeq`. \citet{Hebestreit:2013ko} changed the default minimum window size to $80$ bp but still required at least $20$ CpGs per window. \citet{Lacey:2013iy} kept the default minimum window size of $2,000$ bp but reduced the minimum number of CpGs per window to $50$ from the default of $70$. Nevertheless, the fact remains that `bsseq` is designed for analysing whole-genome bisulfite-sequencing data and not RRBS, which puts it at a disadvantage in these comparisons.

@@

Under the (conditional) binomial model, $M_{i, j} | d_{i, j} \eqd Bin(d_{i, j}, B_{i, j})$, $\beta_{i, j} = \frac{m_{i, j}}{m_{i, j} + u_{i, j}}$ is an unbiased estimator of $B_{i, j}$ with standard error $se(\beta_{i, j}) = \sqrt{\frac{\beta_{i, j}(1 - \beta_{i, j})}{d_{i, j}}}$ \citep{Hansen:2012gr}. The natural interpretation of $\beta_{i, j}$ is then as an estimator of the average level of methylation at the $i^{th}$ locus in the $j^{th}$ sample. In this section I discuss this interpretation and statistical properties of this estimator.

@@

What is immediately clear is that these distributions are bimodal: most CpGs are highly methylated or lowly methylated. The exception is the _E16.5\_male\_1_ sample from the _Seisenberger_ data which is hypomethylated and with an enormous number of CpGs that have intermediate methylation levels. The _E16.5\_male\_1_ sample is progenitor germ cells from a pool of embryonic day 16.5 male mice. Between days E6.5 and E13.5, the mouse progenitor germ cells undergo global demethylation and it is only from day E16.5 onwards that they begin to be _de novo_ methylated \citep{Seisenberger:2012ko}, hence the wide variation in $\beta$-values.

@@

These plots show that CpG islands have a more strictly bimodal distribution than do the non-islands. While the majority of CpGs in CpG islands are unmethylated, there are a subset of methylated CpGs in CpG islands in each sample (except for the _E16.5\_male\_1_ sample). The _H1\_r1_ and _H1\_r2_ samples, replicates of an embryonic stem cell line, stand out for having CpGs in CpG islands being more methylated than unmethylated. These plots also show that most of the intermediate methylation occurs outside of the CpG islands.

@@

Some of the embryonic stem cell samples (_H1\_r1_, _H1\_r2_ and _HSF1_ from the _Lister_ dataset) have less correlated strand-specific $\beta$-values, $r = 0.5$ to $0.7$. The other embryonic stem cell samples have higher correlations of $\beta$-values across strands, although the estimates are quite different between the two replicates of the same cell line (_H9_: $r = 0.77$, _H9\_Laurent_: $r = 0.92$). This suggests that caution may be warranted in combining CpG methylation levels across strands for embryonic stem cell samples.

@@

The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine. Note that this refers to individual cytosines, $Z_{h, i}$, and not the genomic position of the locus, $pos_{i}$. Also recall that most assays measure DNA methylation from a pool of cells, not a single haploid copy of the genome. The methylation state at the $i^{th}$ locus may vary across the $H_{i}$ DNA fragments. Therefore, I do not think it makes sense to describe a locus, $i$, as being a 'methylcytosine'. However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have used this latter definition, which I believe to be an unnecessary source of confusion.

@@

Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, nor is any code made available, I believe the method is as follows[^mathematical_description]. For each cytosine they compute the probability of observing more than $m_{i}$ methylated reads by chance, $P_{i} = \sum_{k = m_{i} + 1}^{k = d_{i}} Pr(X = k)$, where $X = Binom(d_{i}, \epsilon)$ and $\epsilon$ is the estimated 'error'. Any site with an FDR-adjusted P-value below a threshold was declared a "methylcytosine"[^fdr]

@@

\citet{Du:2010dc} advocate for the use of $\mathcal{M}$-values for conducting differential methylation analysis from microarray data[^interpretation]. The main reason they advocate for the use of $\mathcal{M}$-values is that they are approximately _homoscedastic_, i.e. their variances are approximately constant across the full range of $\mathcal{M}$-values. As already noted, the logit-transformation is not the only possible variance-stabilising transformation, but the familiarity of log-ratios to bioinformaticians and genomics researchers makes it a favourable choice. As with $\beta$-values, the $\mathcal{M}$-values derived from bisulfite-sequencing data cannot generally be directly analysed due to the variable sequencing coverage across loci.

@@

@@

This chapter has defined a mathematical framework for describing data from whole-genome bisulfite-sequencing data. It has addressed some subtleties and complications that arise due to within-sample and between-sample differences in where DNA methylation is measured. By using a common statistical framework we can better understand how different statistical methods relate to one another. Using this framework, we described common estimators of DNA methylation levels and some of their statistical properties. We then examined the empirical distributions of these variables across a diverse set of 40 whole-genome bisulfite-sequencing samples.

From these analyses we have seen that the CpG islands drive the strong bimodal distribution of $\beta$-values that is observed in almost all samples. We have also observed that most intermediate methylation occurs outside of the CpG islands. The most distinct methylomes are the hypermethylated embryonic stem cells (_H1\_r1_ and _H1\_r2_) and the hypomethylated cancer cell lines (_HepG2\_cell\_line_). The Seisenberger samples also stand out, particularly the hypervariable progenitor germ cells (_E16.5\_male\_1_). The genome-level results for the _Seisenberger_ samples are more difficult to interpret, however, since they are from pooled DNA. Nonetheless, some of the increased variability in the _Seisenberger_ data will also reflect that these samples are from developmental timepoints during which DNA methylation is very dynamic.
